title: The Effects of Unequal Healthcare Resource Distribution on Life Expectancy: A Data Driven Approach
author: Sunghee Park, Amber Wu, Teerth Mehta, Alex Cao
date: 11/14/22
output: html_document

---

# The Effects of Unequal Healthcare Resource Distribution on Life Expectancy: A Data Driven Approach
#### Author: Sunghee Park, Amber Wu, Teerth Mehta, Alex Cao
#### Date: 11/14/22

## Introduction


## Summary


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, include = FALSE}
# Example 1: Note relative path, which can be read: Up one
# directory(..), down into source (/source), and 
# then "source" an R file (data_access.R)
source("../source/aggregate_table_script_b.3.R")
source("../source/medical_debt_analysis_b4.R")
# data_access_test()
```

## Table

&nbsp;
&nbsp;


<center> <h3> __Top Ten Counties in the US with Highest Percentage of Medical Debt in 2022__ <h3> </center>


&nbsp;

```{r, echo = FALSE}
# Example 1: This function was "sourced" above
# msg <- data_access_test(" Morgan!")
medical_debt_table_10 <- medical_debt_DT
knitr::kable(medical_debt_table_10, format = "pipe")
```
This table is grouped by the ten highest percentage of medical debt. Greene County, a non-metropolitan county, in North Carolina has the highest share with medical debt of 43.8% in the United States. The table shows that 9 out of 10 counties with highest percentage of medical debt are non-metropolitan counties. We also noticed that all of the counties in the table are located in Southern or Eastern United States.
  
&nbsp;
&nbsp;

## Chart 1: Average Medical Debt in Metro and Non-Metro Areas
```{r, echo = FALSE}
disp_med_debt
```

